Notes about Spin and TR Symmetry

### 2016-10-22 Notes of Bernvig's Book/Ch4 TR symmetry/Notes about TR symmetry.tex
New notes about TR symmetry

    We then have the useful fact:
    \begin{fact}
        For any kets $\ket{\alpha},\ket{\beta}$, we have
        \begin{align}
            \pm \braket{\alpha|T\beta} = 
                \braket{\beta|T\alpha}
        \end{align}
        The sign is the same as in $T^2=\pm 1$. Specifically, one has
        \begin{align}
            \braket{\alpha |T \alpha} = 0
        \end{align}
        when $T^2= -\mathbbm{1}$.
    \end{fact}
    \begin{proof}
        This is simple by using 
        $\braket{T\alpha|T\beta} = \braket{\beta|\alpha}$, 
        and replacing $\ket{\alpha}$ with $\ket{T \alpha}$.
    \end{proof}

@@

            Here 
            $$F_{ij}(\vb{k}) = -i \left( \braket{\partial_i
            u(k)|\partial_j u(k)} - (i\leftrightarrow j) \right)$$
            is the Berry field strength.
            For system with TR symmetry, we have
            $$ u(k) = T u(-k)$$
            (Note that $T$ flips the sign of $k$) Since $T$ is
            antiunitary, we have $ u(-k) = u(k)^*$. Therefore
            $\braket{\partial_i u(-k)|\partial_j u(-k)} = 
              \braket{\partial_j u(k)|\partial_i u(k)}$
            So:
            \begin{equation}
                F_{ij}(\vb{-k}) = - F_{ij}(\vb{k})
            \end{equation}
            Therefore $F_{ij}$ is a odd function. So its integration
            over the whole BZ is zero, i.e. $\sigma_{ij}$ is zero.
        \end{proof}

>>>>

    We then have the useful fact:
    \begin{fact}
        \label{fact:TROperator_TRInSpin_InnerProdTUandU}
        For any kets $\ket{\alpha},\ket{\beta}$, we have
        \begin{align}
            \pm \braket{\alpha|T\beta} = 
                \braket{\beta|T\alpha}
        \end{align}
        The sign is the same as in $T^2=\pm 1$. Specifically, one has
        \begin{align}
            \braket{\alpha |T \alpha} = 0
        \end{align}
        when $T^2= -\mathbbm{1}$.
    \end{fact}
    \begin{proof}
        This is simple by using 
        $\braket{T\alpha|T\beta} = \braket{\beta|\alpha}$, 
        and replacing $\ket{\alpha}$ with $\ket{T \alpha}$.
    \end{proof}

@@

            Here 
            $$F_{ij}(\vb{k}) = -i \left( \braket{\partial_i
            u(k)|\partial_j u(k)} - (i\leftrightarrow j) \right)$$
            is the Berry field strength.
            For system with TR symmetry, we have
            $$ u(k) = T u(-k)$$
            (Note that $T$ flips the sign of $k$) Since $T$ is
            antiunitary, we have $ u(-k) = u(k)^*$. Therefore
            $\braket{\partial_i u(-k)|\partial_j u(-k)} = 
              \braket{\partial_j u(k)|\partial_i u(k)}$
            So:
            \begin{equation}
                F_{ij}(\vb{-k}) = - F_{ij}(\vb{k})
            \end{equation}
            Therefore $F_{ij}$ is a odd function. So its integration
            over the whole BZ is zero, i.e. $\sigma_{ij}$ is zero.
        \end{proof}
        \subsubsection{Time reversal symmetry in electron bands}
        \label{sec:Time-reversal-symmetry-in-electron-bands}
        Natually operator $T$ flips the sign of the
        creation/annihilation operator of an electron. We now proves:
        \begin{thm}
            We have, up to a phase,
            \begin{align}
                & T c_\sigma T^{-1} =
                i(\sigma_y)_{\sigma\sigma'}c_{\sigma'} \\
                & T c_\sigma^\dagger T^{-1} = 
                ic_{\sigma'}^\dagger(\sigma_y)^T_{\sigma'\sigma}
            \end{align}
            Where $\sigma_y = \left( \begin{array}{cc}
                         0 & -i \\
                         i & 0 \\ \end{array} \right)$.
            and $\sigma$ label spin $\uparrow$ and spin
            $\downarrow$.
        \end{thm}
        \begin{proof}
            Assuming
            \begin{align}
                & Tc_\uparrow T^{-1} = A c_\downarrow \\
                & Tc_\downarrow T^{-1} = B c_\uparrow
            \end{align}
            Then we need only to prove $AB^* = -1$. Since up to a
            phase change, we can let
            \begin{equation}
                A = 1\,, B = -1
            \end{equation}
            This is what the theorem says.

            Let's consider the following
            $$T c_\uparrow c_\uparrow^\dagger \ket{0}$$
            First, it is equal to $T\ket{0}$. On the other hand, we
            can let the $T$ jumps over one creation operator:
            $$A c_\downarrow T c_\uparrow^\dagger \ket{0}$$
            Now insert $T^{-1}T$ between $A$ and $c_\downarrow$. And
            makes the new $T$ one jump to the right:\
            $$ A T^{-1} B c_\uparrow T^2 c_\uparrow^\dagger \ket{0}$$
            Now be careful to notice that the first $T^{-1}$ is $T$,
            whereas the second $T^2$ is $- \mathbbm{1}$ (watch the
            Spin!). With simple calculation one finds it is equal to
            $$ -AB^* T\ket{0}$$
            Hence $-AB^* = 1$. The case for the creation operatros is
            similar.
        \end{proof}
        \begin{thm}
            For the creation and annihilation in the momentum space,
            we have:
            \begin{align}
                & Tc^\dagger_{\vb{k}\sigma} T^{-1} =
                    c^\dagger_{\vb{-k}\sigma'}
                    i(\sigma_y)^T_{\sigma\sigma'} \\
                & Tc_{\vb{k}\sigma} T^{-1} =
                    c_{\vb{-k}\sigma'}
                    i(\sigma_y)_{\sigma\sigma'}
            \end{align}
        \end{thm}
        \begin{proof}
            This is quite obvious since $T$ changes the coefficient of
            fourier expansion: $e^{-i \vb{k}\vdot\vb{R}_j} \to 
            e^{i \vb{k}\vdot\vb{R}_j}$
        \end{proof}
        Then directly apply $T$ and $T^{-1}$ to both sides of
        $$H = \sum_{\vb{k}} c^\dagger_{\vb{k}\sigma}
            h^{\sigma\sigma'}(\vb{k})
            c_{\vb{k}\sigma'}$$

            gives (noticing $h(k)$ is just a number that $T$ turns it
            into its conjugate number):
        \begin{equation}
            h^{\sigma\sigma'}(\vb{k}) \mapsto
            i(\sigma_y)^T_{\sigma_1\sigma} (h^{\sigma,\sigma'})^*
            i(\sigma_y)_{\sigma'\sigma_2}
        \end{equation}
        If one carefully examine the cases, one finds that when
        $\sigma_1\neq\sigma_2$, the above should obviously be $0$. And
        when $\sigma_1=\sigma_2$, the RHS above is surprisingly just
        $(h^{\{\cdot,\cdot\}})^*$, with appropriate spin.
        So we have compactly:
        \begin{thm}
            For electron bands,
            \begin{equation}
                Th(\vb{k})T^{-1} = h(-\vb{k})
            \end{equation}
            with spin inverted.
        \end{thm}

        Then we have the following fact:
        \begin{fact}
            For Time reversal symmetric electron bands, there is
            always at least two degenerate state
        \end{fact}
        This is obvious since:
        $$h(-\vb{k}) T\ket{u(\vb{k})} = Th(\vb{k})\ket{u(\vb{k})}
        = TE(\vb{k})\ket{u(\vb{k})} = E(\vb{k})T\ket{u(\vb{k})} $$
        Also, by fact \ref{fact:TROperator_TRInSpin_InnerProdTUandU},
        one sees that the time reversed state $T\ket{u(\vb{k})}$ is
        distinct from the orginal state. Hence the double degeneracy.
